---
title: "RNAseq cell type deconvolution"
output: html_document
---

# Prepare data for CIBERSORTx
### Setup
I'm using CIBERSORTx for cell deconvolution.

Brain signature matrix is from ROSMAP data, as published in Dai et al., 2024 (https://www-science-org.dartmouth.idm.oclc.org/doi/10.1126/sciadv.adh2588). 
The dataset contains 8 cell types including Excitatory neurons, Oligodendrocytes, Inhibitory neurons, Microglia, Oligodendrocyte precursor cells, Astrocytes, Endothelial cells and Pericytes.
It is CPM (counts per million) normalised.

CIBERSORTx prefers matched normalisation between signature matrix and mixture. I'm reading in raw counts and CPM normalising & log2 transforming.
```{r}
library(readr)
library(glue)
library(edgeR)
library(dplyr)

project_dir <- "~/Documents/repos/explant-seq"
rna_dir <- glue("{project_dir}/data/RNA/RNA-preprocessed")

# read in raw counts
raw_counts <- read.table(
  glue("{rna_dir}/260601_exp001-007/feature_counts/featurecounts.readcounts.raw.tsv"),
  sep = "\t", header = TRUE)

# clean column names & remove unnecessary columns
colnames(raw_counts) <- gsub("alignment.(.*?).srt.bam", "\\1", colnames(raw_counts))
cols <- c(
  'Gene Name', 'Chr', 'Start', 'End', 'Strand', 'Length', 
  'EXP002_Frontal_lobe', 'EXP005_LPAG', 'EXP005_LPRS', 'EXP005_RLTS')
raw_counts <- raw_counts[, !(names(raw_counts) %in% cols)]

# CPM normalise & log2 transform
cpm_log <- raw_counts %>% dplyr::select(-"Geneid") %>% cpm(log = TRUE, prior.count = 1)
cpm_log <- data.frame(cpm_log)
cpm_log$ensembl_gene_id <- raw_counts$Geneid
```

### Get Ensembl ID to gene symbol key 
```{r}
library(biomaRt)

mart <- useMart("ensembl", 
                dataset = "hsapiens_gene_ensembl")

annot <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = cpm_log$ensembl_gene_id,
  mart = mart
)
```

### Clean reference & filter CPM to match reference gene list
```{r}
# merge in gene IDs into cpm_log
cpm_log <- annot %>% left_join(cpm_log, by = "ensembl_gene_id")

# read in ROSMAP, remove MARCH genes (Excel date conversion issue), "versioned" genes and duplicates
rosmap <- read.table(glue("{project_dir}/data/RNA/rosmap_ref_rujiadai_251028.csv"), sep=",", header=TRUE)
rosmap <- rosmap[!grepl("^MARCH", rosmap$X), ]
rosmap$X <- sub("\\..*$", "", rosmap$X)
rosmap <- rosmap[!duplicated(rosmap$X), ]
write.table(rosmap, glue("{project_dir}/data/RNA/cibersort_signature_matrix_rosmap_clean_260608.tsv"), sep = "\t", row.names = FALSE)

# number of gene ids that match between the Rosmap signature matrix and my RNAseq data
matching <- intersect(cpm_log$hgnc_symbol, rosmap$X)
cat("Number of matching gene IDs between ROSMAP & electrode RNAseq data:", length(matching))

# filter by matching gene IDs
filtered <- cpm_log[cpm_log$hgnc_symbol %in% matching, ]
filtered <- filtered[!duplicated(filtered$hgnc_symbol), ]
filtered <- filtered %>% dplyr::select(-"ensembl_gene_id")

write.table(filtered, glue("{project_dir}/data/RNA/cibersort_mixture_cpm_log2_260608.tsv"), sep = "\t", row.names = FALSE)
```

# Analyse deconvoluted cell proportions
```{r}
library(ggplot2)
library(reshape2)

deconv_df <- read.csv(glue("{project_dir}/data/RNA/cibersort_res_job11_260609.csv"))
deconv_df <- deconv_df %>% dplyr::select(-c("P.value", "Correlation", "RMSE"))
deconv_df_melt <- reshape2::melt(deconv_df, id.vars = "Mixture")

ggplot(data=deconv_df_melt, aes(x=variable, y=value, colour = variable)) + geom_boxplot()
```

```{r}
library(ggplot2)
library(wesanderson)

zissou_pal <- wes_palette("Zissou1", 8, type = "continuous")

p_box <- ggplot(deconv_df_melt, aes(x = variable, y = value, colour = variable, fill = variable)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
  scale_colour_manual(values = zissou_pal) +
  scale_fill_manual(values = zissou_pal) +
  labs(
    title = "Overall deconvolved cell proportions (ROSMAP reference)",
    subtitle = "n=52 samples",
    x = NULL,
    y = "Proportion",
    colour = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey50"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p_stack <- ggplot(deconv_df_melt, aes(x = Mixture, y = value, fill = variable)) +
  geom_bar(stat = "identity", width = 1) +
  scale_fill_manual(values = zissou_pal) +
  labs(
    title = "Deconvolved cell proportions across 52 samples (ROSMAP ref)",
    subtitle = "In, Opc, Ast had 0 proportions",
    x = NULL,
    y = "Proportion (%)",
    fill = "Cell type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, colour = "grey50"),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    panel.grid = element_blank(),
    legend.position = "right"
  )


p_box
p_stack
```







