---
title: "DGE hippomcapus analysis"
output: html_document
---

# Setup

```{r}
library(readr)
library(glue)
library(dplyr)
library(purrr)
library(readxl)


project_dir <- "~/Documents/repos/explant-seq"
rna_dir <- glue("{project_dir}/data/RNA/RNA-preprocessed")
patients <- c("EXP001", "EXP003", "EXP006", "EXP007")

# read in electrode metadata
electrode_metadata <- patients %>%
  set_names() %>%
  map_dfr(~ read_excel(glue("{project_dir}/data/metadata/{.x}_electrode_metadata.xlsx")),
          .id = "subject")
meta <- electrode_metadata[!is.na(electrode_metadata$hippocampus), ]

# read in raw counts
raw_counts <- read.table(
  glue("{rna_dir}/260601_exp001-007/feature_counts/featurecounts.readcounts.raw.tsv"),
  sep = "\t", header = TRUE)
colnames(raw_counts) <- gsub("alignment.(.*?).srt.bam", "\\1", colnames(raw_counts))

cols <- c(meta$sample_id, "Geneid")
raw_counts <- raw_counts[, (names(raw_counts) %in% cols)]
```

```{r}
library(DESeq2)
library(tibble)

# ensure electrode metadata rows match raw_counts
raw_counts <- column_to_rownames(raw_counts, var = "Geneid")
meta <- meta[match(colnames(raw_counts), meta$sample_id), ]
all(colnames(raw_counts) == meta$sample_id)
meta$hippocampus <- factor(meta$hippocampus, levels = c("contralateral", "implicated")) # contralateral as baseline)
meta$subject <- factor(meta$subject)

# create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = raw_counts,
  colData = meta,
  design = ~ subject + hippocampus
)

# filter to remove genes with very low counts (keeps genes with >10 counts in >4 samples)
keep <- rowSums(counts(dds) >= 10) >= 4
dds <- dds[keep, ]

# run DESeq2
dds <- DESeq(dds)
resultsNames(dds)
res <- results(dds, name = "hippocampus_implicated_vs_contralateral", alpha = 0.1)
summary(res)
res_df <- as.data.frame(res) %>%
  rownames_to_column("ensembl_gene_id") %>%
  arrange(padj)

head(res_df)
```

```{r}
library(biomaRt)

mart <- useMart("ensembl", 
                dataset = "hsapiens_gene_ensembl")

annot <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = res_df$ensembl_gene_id,
  mart = mart
)

res_df <- annot %>% left_join(res_df, by = "ensembl_gene_id")
res_df <- res_df[order(res_df$pvalue), ]
```

```{r}
library(ggplot2)
library(ggrepel)

# define significance thresholds
lfc_thresh <- 1
p_thresh <- 0.001

# categorise genes
res_df <- res_df %>%
  mutate(
    significance = case_when(pvalue < p_thresh & log2FoldChange > lfc_thresh  ~ "Up",
                             pvalue < p_thresh & log2FoldChange < -lfc_thresh ~ "Down" , 
                             TRUE ~ "NS"),
    significance = factor(significance, levels = c("Up", "Down", "NS"))) %>%
  filter(!is.na(hgnc_symbol) & hgnc_symbol != "") %>%
  filter(!is.na(pvalue)) %>%
  mutate(fc_pval_rank = sign(log2FoldChange) * -log10(pvalue)) %>%
  arrange(desc(fc_pval_rank))


# select top genes to label
label_genes <- res_df %>%
  filter(pvalue < p_thresh) %>%
  arrange(desc(abs(log2FoldChange))) %>%
  slice_head(n = 10)

ggplot(res_df, aes(x = log2FoldChange, y = -log10(pvalue), colour = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_point(data = filter(res_df, significance != "NS"), size = 2.5, alpha = 0.9) +
  geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), 
             linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_hline(yintercept = -log10(max(res_df$pvalue[res_df$pvalue < p_thresh], na.rm = TRUE)),
             linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_text_repel(
    data = label_genes,
    aes(label = hgnc_symbol),
    size = 3.5,
    max.overlaps = 20,
    box.padding = 0.5,
    colour = "black") +
  scale_colour_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5", "NS" = "grey70")) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(title = "Contralateral vs Implicated Hippocampus", x = "log2FC", y = "-log10(pval)", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

# GSEA
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(dplyr)

head(res_df)

ranked_list <- res_df %>%
  filter(!is.na(fc_pval_rank)) %>%
  arrange(desc(fc_pval_rank)) %>%
  # convert HGNC to Entrez ID
  left_join(
    bitr(res_df$hgnc_symbol, 
         fromType = "SYMBOL",
         toType = "ENTREZID",
         OrgDb = org.Hs.eg.db),
    by = c("hgnc_symbol" = "SYMBOL")
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  # remove duplicate entrez IDs keeping highest absolute rank
  arrange(desc(abs(fc_pval_rank))) %>%
  filter(!duplicated(ENTREZID)) %>%
  # create named vector for GSEA
  { setNames(.$fc_pval_rank, .$ENTREZID) } %>%
  sort(decreasing = TRUE)

gsea_go <- gseGO(
  geneList     = ranked_list,
  OrgDb        = org.Hs.eg.db,
  ont          = "BP",
  minGSSize    = 15,
  maxGSSize    = 500,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  verbose      = TRUE,
  seed         = 42
)

# simplify redundant GO terms (cutoff = similarity threshold)
gsea_go_simplified <- simplify(gsea_go, cutoff = 0.5, by = "p.adjust", select_fun = min)
as.data.frame(gsea_go_simplified)
```

```{r}
df <- as.data.frame(gsea_go_simplified) %>%
  arrange(NES) %>%
  mutate(
    Description = factor(Description, levels = Description),
    Direction   = ifelse(NES > 0, "Enriched in SOZ HA", "Suppressed in SOZ HA"),
    Direction   = factor(Direction, levels = c("Suppressed in SOZ HA", "Enriched in SOZ HA"))
  )

# Plot
ggplot(df, aes(x = NES, y = Description, fill = Direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Suppressed in SOZ HA" = "#4647AE", "Enriched in SOZ HA" = "#34908B")) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = "black") +
  labs(
    title = "GSEA GO Biological Process",
    x     = "Normalised Enrichment Score (NES)",
    y     = NULL,
    fill  = NULL
  ) +
  theme_classic() +
  theme(
    axis.text.y  = element_text(size = 8),
    legend.position = "bottom"
  )
```

Implicated hippocampi were enriched in pathways related to cholesterol biosynthesis (higher cholesterol can contribute to enhanced glutamatergic transmission), skeletal system morphogenesis and defence to virus (see leading edge genes below).

```{r}
print_leading_genes <- function(go_id){
  leading_genes <- gsea_go@result[go_id, "core_enrichment"]
  gene_info <- bitr(
    strsplit(leading_genes, split = "/")[[1]],
    fromType = "ENTREZID",
    toType   = c("SYMBOL", "GENENAME"),
    OrgDb    = org.Hs.eg.db
  )
  print(gene_info)
}

# 0048705 skeletal system morphogenesis
print_leading_genes("GO:0048705")
```
Skeletal system morphogenesis really relates to extracellular matrix remodelling genes, that includes matrix metallopeptidases, collagens, hyaluronidase etc. Interesting is histone lysine methyltransferase which is an epigenetic regulator - the ECM changes might be more long-term than just acute. MEF2C is an important regulator of synaptic pruning.

```{r}
# 0051607 defence response to virus
print_leading_genes("GO:0051607")
```
This GO term might reflect increased sterile neuroinflammation in the implicated hippocampus rather than defence response to virus:
- damaged/stressed cells release damage-associated molecular patterns (DAMPs)
- these are recognised by innate immune sensors such as cGAS, TLR3 (double stranded RNA), ZBP1 (left-handed DNA, RNA)
- this then results in type I interferon response (STAT1, STAT2, IRF1, IFNAR2, IKBKE)
- TRIM family ubiquitin ligases are interferon-stimulated genes





