---
title: "RNAseq exploratory analysis"
output: html_document
---

# QC: mapped and assigned reads
HISAT2 alignment score relates to the number of reads that were mapped to the reference genome.
Feature counts QC summarises the quantity of those mapped reads that overlap with an annotated feature on the reference genome.

### Setup
```{r}
library(readr)
library(glue)
library(janitor)
library(readxl)
library(dplyr)
library(purrr)

project_dir <- "~/Documents/repos/explant-seq"
rna_dir <- glue("{project_dir}/data/RNA/RNA-preprocessed")
patients <- c("EXP001", "EXP003", "EXP006", "EXP007")
```

### Extract proportions of mapped reads from alignment txt summaries
```{r}
parse_hisat_summary <- function(filepath) {
  lines <- readLines(filepath)
  rate_line <- lines[grepl("overall alignment rate", lines)]
  rate <- as.numeric(gsub("%.*", "", rate_line))
  tibble(file = basename(filepath), perc_mapped = rate)
}

alignment_qc <- patients %>%
  set_names() %>%
  map_dfr(~ {
    files <- Sys.glob(glue("{rna_dir}/260601_exp001-007/alignment_summary/{.x}_*.hisat.summary.txt"))
    map_dfr(files, parse_hisat_summary)
  }, .id = "subject") %>%
  mutate(sample = gsub("\\.hisat\\.summary\\.txt", "", file)) %>% 
  select(subject, sample, perc_mapped)
```

### Extract proportions of assigned reads from featurecounts summary file
```{r}
fc_summary <- read_tsv(glue("{rna_dir}/260601_exp001-007/feature_counts/featurecounts.readcounts.raw.tsv.summary")) %>%
  as.data.frame() %>%
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  tibble::rownames_to_column("sample") %>%
  mutate(sample = gsub("alignment/(.*?)\\.srt\\.bam", "\\1", sample)) %>%
  mutate(across(-sample, as.numeric)) %>%
  mutate(total = rowSums(across(-sample)),
         perc_assigned = (Assigned / total) * 100) %>%
  select(sample, perc_assigned)

qc_data <- left_join(alignment_qc, fc_summary, by = "sample")
```

```{r}
library(ggplot2)
library(tidyr)

qc_data %>%
  pivot_longer(cols = c(perc_mapped, perc_assigned),
               names_to = "metric",
               values_to = "value") %>%
  mutate(metric = recode(metric,
                         "perc_mapped" = "% Mapped",
                         "perc_assigned" = "% Assigned"), 
         metric = factor(metric, levels = c("% Mapped", "% Assigned"))) %>%
  ggplot(aes(x = metric, y = value, colour = subject)) +
  geom_boxplot(outlier.shape = NA, colour = "grey60", width = 0.4) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  labs(title = "Alignment and assignment rates 4 patients (52 electrodes)",
       x = NULL, y = "%", colour = "Subject") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank())
```

# EXPLORATORY ANALYSIS (PCA, batch correction)
### Setup
Principal component analysis is done on the TPM (transcripts per million) normalised data. However, 
```{r}
# read in electrode metadata
electrode_metadata <- patients %>%
  set_names() %>%
  map_dfr(~ read_excel(glue("{project_dir}/data/metadata/{.x}_electrode_metadata.xlsx")),
          .id = "subject")

# read in TPM normalised counts
tpm_counts <- read_tsv(
  glue("{rna_dir}/260601_exp001-007/feature_counts/featurecounts.readcounts_tpm.ann.tsv"))
cols <- c(
  'Gene Name', 'Chr', 'Start', 'End', 'Strand', 'Length', 
  'EXP002_Frontal_lobe', 'EXP005_LPAG', 'EXP005_LPRS', 'EXP005_RLTS')
tpm_counts <- tpm_counts[, !(names(tpm_counts) %in% cols)]
```

### Filter and log transform
```{r}
library(edgeR)

# remove rows where more than 90% of values are zero
tpm_counts <- tibble::column_to_rownames(tpm_counts, var = "Ensembl ID")
zero_fraction <- rowSums(tpm_counts == 0) / ncol(tpm_counts)
tpm_matrix <- data.matrix(tpm_counts[zero_fraction <= 0.9, ])  # filter and convert to data matrix
# add a small pseudocount to avoid log(0)
tpm_log <- log2(tpm_matrix + 1)
head(tpm_log)
```

### PCA before batch correction
```{r}
# ggplot theme for PCA plots
theme_prism <- theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),          # remove all panel borders
    axis.line = element_line(color = "black", linewidth = 1),  # draws bottom/left axes
    axis.ticks = element_line(color = "black", linewidth = 1),
    axis.title.y = element_text(face = "bold", size = 14),
    axis.title.x = element_text(face = "bold", size = 14),
    axis.text = element_text(face = "bold", size = 12, color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    plot.margin = margin(10, 10, 10, 10)
  )

# PCA before batch correction
pca_no_bc <- prcomp(t(tpm_log), scale. = TRUE)
# extract PC1 and PC2 and merge in metadata
pca_no_bc_df <- data.frame(
  PC1 = pca_no_bc$x[,1],
  PC2 = pca_no_bc$x[,2],
  electrode_metadata
)

var_explained <- pca_no_bc$sdev^2 / sum(pca_no_bc$sdev^2)
p_no_bc <- ggplot(pca_no_bc_df, aes(x = PC1, y = PC2, color = subject)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "PCA before batch correction",
    x = paste0("PC1 (", round(var_explained[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2]*100, 1), "%)")
  ) + theme_prism +
  scale_color_brewer(palette = "Dark2")
```

### Batch correction using SVA's ComBat & Save data
```{r}
library(sva)

electrode_metadata <- electrode_metadata[match(colnames(tpm_log), electrode_metadata$sample_id), ]
all(colnames(tpm_log) == electrode_metadata$sample_id)

combat_expr <- ComBat(
  dat = tpm_log,
  batch = electrode_metadata$subject,
  par.prior = TRUE,
  prior.plots = FALSE
)
# write.table(combat_expr, glue("{project_dir}/data/RNA/RNA-analysis/log2transformed_batch_corrected_gene_exp_mat.csv"), sep = "\t")
```


### Batch correction using SVA's ComBat & Save data
```{r}

```



```{r}
pca_bc <- prcomp(t(combat_expr), scale. = TRUE)
pca_bc_df <- data.frame(
  PC1 = pca_bc$x[,1],
  PC2 = pca_bc$x[,2],
  electrode_metadata
)

var_explained <- pca_bc$sdev^2 / sum(pca_bc$sdev^2)
p_bc <- ggplot(pca_bc_df, aes(x = PC1, y = PC2, color = subject)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "PCA after batch correction",
    x = paste0("PC1 (", round(var_explained[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2]*100, 1), "%)")
  ) + theme_prism +
  scale_color_brewer(palette = "Dark2")
```

```{r}
p_no_bc
p_bc
```

PCs capture less variance than prior to the batch correction (PC1: 13.7% -> 6.5%, PC2: 4.2% -> 4.1%). However, Combat did not adjust away the difference between EXP003 and the other subjects. This might be due to true biological differences between subjects. We could also explore more aggressive batch correction approaches.

```{r}
p <- ggplot(pca_bc_df[pca_bc_df$region != "x", ], aes(x = PC1, y = PC2, color = region)) +
  geom_point(size = 3) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "PCA coloured by brain region",
    x = paste0("PC1 (", round(var_explained[1]*100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2]*100, 1), "%)")
  ) +
  theme_prism
p
```




